Understanding the role of different proxy selection methods in High-Dimensional Propensity Score Analysis

### 2024-09-02 Writeup/manuscript.Rmd
app6

```{r, echo=FALSE, out.width="100%", fig.cap="Comparison of Bias Across Different Methods in hdPS Analysis", fig.align="center", fig.label="fig:bias-comparison"}
knitr::include_graphics("figures/metric_comparison_Bias.png")
```

```{r, echo=FALSE, out.width="100%", fig.cap="Comparison of Coverage Probability Across Different Methods in hdPS Analysis", fig.align="center", fig.label="fig:bias-comparison"}
knitr::include_graphics("figures/metric_comparison_Coverage.png")
```

@@

```{r, echo=FALSE, fig.align="center", fig.height=6, fig.width=8, fig.cap="Figure presenting a comparison of Risk Differences (RD) and Odds Ratios (OR) with 95% confidence intervals for different methods used to evaluate the association between obesity and diabetes risk. The analysis is based on data from the National Health and Nutrition Examination Survey (NHANES) for the years 2013-2018. Methods are arranged by the number of variables used in the models.", fig.label="fig:comparison-plot", message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)

@@

@@

1. MSE - Mean Squared Error
2. SE - Standard Error
3. PS - Propensity Score
4. AE - Autoencoders
5. DL - Deep Learning
6. MARS - Multivariate Adaptive Regression Splines
7. SMD - Standardized Mean Difference
8. TMLE - Targeted Maximum Likelihood Estimation
9. RHC - Right Heart Catheterization
10. SUPPORT - Study to Understand Prognoses and Preferences for Outcomes and Risks of Treatments

>>>>

The results for each method under the different scenarios are summarized below. See Figures \ref{fig:bias-comparison} and \ref{fig:Coverage-comparison} for an overview of the performance in terms of bias and coverage, respectively.

(i) **Frequent Exposure and Outcome (base) scenario**:

1. *Bias*: The kitchen sink model, which includes all variables without selection, exhibited the smallest bias (0.0002). In contrast, the Genetic Algorithm (GA) showed the highest bias (0.0287). Among the other methods, Bross (-0.0001), Hybrid (0.0016), and Elasticnet (0.0036) demonstrated low bias, indicating that these methods provide estimates closer to the true effect. XGBoost (0.0074) and Random Forest (RF) (0.0034) had slightly higher bias but remained within acceptable limits.

2. *Coverage*: Most methods, including Hybrid, Forward, Backward, LASSO, and Elasticnet, achieved high coverage values around 98$\%$, indicating well-calibrated confidence intervals. However, the GA method had significantly lower coverage (83.8$\%$), suggesting that its confidence intervals might be too narrow or biased, potentially missing the true effect. Given the bias in the GA method's results, we also calculated bias-eliminated coverage. This adjustment improved GA’s coverage to 96$\%$, but it still remained lower than that of the other methods.

3. *Mean Squared Error (MSE)*: XGBoost achieved the lowest MSE (0.0006), indicating it as the most accurate method overall. Conversely, the GA method had the highest MSE (0.0016), reflecting its higher bias and variability. The kitchen sink model (0.0009), Bross (0.0008), Hybrid (0.0008), and Elasticnet (0.0009) all had relatively similar and moderate MSE values.

4. *Standard Error (SE)*: The lowest Empirical SE was observed with XGBoost (0.0229), indicating high precision in its estimates. The kitchen sink model had the highest Empirical SE (0.0305), suggesting greater variability. Other methods such as GA (0.0274), Hybrid (0.0278), and Bross (0.0287) exhibited moderate variability, while LASSO (0.0299) and Elasticnet (0.0294) had slightly higher variability. Similarly, XGBoost also showed the lowest Model-based SE (0.0268), consistent with its low Empirical SE and indicating high precision. The kitchen sink model had the highest Model-based SE (0.0333), suggesting less precision in its estimates.

**(ii) Rare Exposure and Frequent Outcome**:

1. *Bias*: In this scenario, the kitchen sink model showed a relatively low bias (0.0025), but the Genetic Algorithm (GA) had the highest bias (0.0408), indicating a substantial deviation from the true effect. XGBoost (0.0259) also exhibited higher bias compared to other methods. On the other hand, Bross (0.0035), Hybrid (0.0049), and Elasticnet (0.0053) demonstrated moderate bias, while Random Forest (RF) (0.0127) and Forward Selection (0.0108) had slightly higher bias but remained within an acceptable range.

2. *Coverage*: Most methods maintained high coverage levels, with XGBoost showing the highest coverage (96.2$\%$), suggesting that its confidence intervals were well-calibrated despite the higher bias. The GA method, however, had slightly lower coverage (92.2$\%$), indicating that its confidence intervals might be narrower, potentially excluding the true effect. Other methods such as RF, Forward, Backward, and Hybrid had coverage values around 95$\%$, suggesting adequate interval calibration. To account for the bias in GA, bias-eliminated coverage was calculated, which improved the coverage for GA to 94.2$\%$, still slightly lower than other methods.

3. *Mean Squared Error (MSE)*: XGBoost and Hybrid methods both demonstrated the lowest MSE (0.0032), indicating that these methods were the most accurate in this scenario. The GA method, with a higher MSE (0.0043), reflected its substantial bias and variability. The kitchen sink model had an MSE of 0.0043, similar to GA, while other methods such as Bross, RF, and Elasticnet exhibited moderate MSE values, indicating reasonable accuracy.

4. *Standard Error (SE)*: The lowest Empirical SE was observed with XGBoost (0.0507) and GA (0.0510), reflecting the high precision of these methods despite the higher bias. The kitchen sink model exhibited the highest Empirical SE (0.0656), indicating greater variability. Methods like Hybrid (0.0564), Bross (0.0595), and RF (0.0609) showed moderate variability, while Elasticnet (0.0597) and LASSO (0.0587) had slightly higher variability. In terms of Model-based SE, XGBoost (0.0531) and GA (0.0533) also showed low variability, while the kitchen sink model had the highest Model-based SE (0.0623), suggesting less precision in its estimates.

**(iii) Frequent Exposure and Rare Outcome**:

1. *Bias*: In this scenario, the kitchen sink model exhibited a moderate negative bias (-0.0093), similar to the Bross method (-0.0088). The Genetic Algorithm (GA) showed a significantly higher bias (0.0362), indicating a substantial deviation from the true effect. Among other methods, XGBoost demonstrated the lowest bias (-0.0061), while methods like Hybrid (-0.0082), Forward (-0.0070), and Backward (-0.0070) had slightly higher but still moderate biases. Elasticnet and LASSO both had biases of -0.0079, reflecting slightly larger deviations compared to XGBoost but still within acceptable limits.

2. *Coverage*: Most methods achieved good coverage, with XGBoost, Random Forest (RF), and Forward Selection each achieving a coverage rate of 95.4$\%$, indicating well-calibrated confidence intervals. The GA method, however, had slightly lower coverage (91.8$\%$), indicating that its confidence intervals might be narrower, potentially excluding the true effect. Bross and the kitchen sink model had slightly lower coverage values of 93.8$\%$ and 93.4$\%$, respectively. After accounting for bias, the bias-eliminated coverage for most methods, except GA, remained high, with values ranging from 98.4$\%$ to 99.0$\%$, indicating that most methods effectively adjusted for bias in their coverage estimates. GA's bias-eliminated coverage was lower at 93.4$\%$, reflecting its higher inherent bias.

3. *Mean Squared Error (MSE)*: XGBoost exhibited the lowest MSE (0.0003), indicating it as the most accurate method overall in this scenario. GA had the highest MSE (0.0040), reflecting its substantial bias and variability. The kitchen sink model (0.0005), Bross (0.0005), and other methods like Hybrid (0.0004) and Elasticnet (0.0005) all had relatively similar MSE values, indicating moderate accuracy.

4. *Standard Error (SE)*: The lowest Empirical SE was observed with XGBoost (0.0152), reflecting high precision in its estimates. The GA method exhibited the highest Empirical SE (0.0523), indicating greater variability and less precision. Methods like Hybrid (0.0184), Forward (0.0187), and Elasticnet (0.0203) showed moderate variability, while Bross (0.0206) and the kitchen sink model (0.0212) had slightly higher variability. In terms of Model-based SE, XGBoost (0.0179) again showed the lowest variability, consistent with its low Empirical SE, indicating that it provided the most stable estimates. The kitchen sink model had a slightly higher Model-based SE (0.0219), indicating less precision in its estimates.

```{r, echo=FALSE, out.width="100%", fig.cap="Comparison of Bias Across Different Methods in hdPS Analysis\\label{fig:bias-comparison}", fig.align="center"}
knitr::include_graphics("figures/metric_comparison_Bias.png")
```

```{r, echo=FALSE, out.width="100%", fig.cap="Comparison of Coverage Probability Across Different Methods in hdPS Analysis\\label{fig:Coverage-comparison}", fig.align="center"}
knitr::include_graphics("figures/metric_comparison_Coverage.png")
```

@@

See Figure \ref{fig:comparison-plot} for the results from analyzing the NHANES (2013-2018) dataset.

```{r, echo=FALSE, fig.align="center", fig.height=6, fig.width=8, fig.cap="Figure presenting a comparison of Risk Differences (RD) and Odds Ratios (OR) with 95% confidence intervals for different methods used to evaluate the association between obesity and diabetes risk. The analysis is based on data from the National Health and Nutrition Examination Survey (NHANES) for the years 2013-2018. Methods are arranged by the number of variables used in the models.\\label{fig:comparison-plot}", message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)

@@

XGBoost consistently outperformed other methods across all scenarios, demonstrating low bias, high coverage, and the lowest MSE. The GA, on the other hand, consistently underperformed with higher bias, lower coverage, and higher MSE, indicating less reliable estimates. The Bross method and the kitchen sink model generally provided moderate performance with low to moderate bias and reasonable coverage, but they were less accurate than XGBoost. Hybrid, Elasticnet, and LASSO methods showed competitive performance with moderate bias, good coverage, and acceptable MSE, making them reliable alternatives, though slightly less optimal than XGBoost.

@@

- hdPS: High-dimensional Propensity Score
- NHANES: National Health and Nutrition Examination Survey
- OR: Odds Ratio
- RD: Risk Difference
- SE: Standard Error
- MSE: Mean Squared Error
- LASSO: Least Absolute Shrinkage and Selection Operator
- GA: Genetic Algorithm
- RF: Random Forest
- XGBoost: Extreme Gradient Boosting
- FS: Forward Selection
- BE: Backward Elimination
- CV: Cross-Validation
- RR: Relative Risk
- Elasticnet: A regularized regression method that combines LASSO and Ridge regression

### 2024-09-02 Writeup/manuscript.tex
app6

@@

\caption{Comparison of Bias Across Different Methods in hdPS Analysis}\label{fig:unnamed-chunk-1}
\end{figure}

@@

\caption{Comparison of Coverage Probability Across Different Methods in hdPS Analysis}\label{fig:unnamed-chunk-2}
\end{figure}

@@

@@

\caption{Figure presenting a comparison of Risk Differences (RD) and Odds Ratios (OR) with 95\% confidence intervals for different methods used to evaluate the association between obesity and diabetes risk. The analysis is based on data from the National Health and Nutrition Examination Survey (NHANES) for the years 2013-2018. Methods are arranged by the number of variables used in the models.}\label{fig:unnamed-chunk-3}
\end{figure}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  MSE - Mean Squared Error
\item
  SE - Standard Error
\item
  PS - Propensity Score
\item
  AE - Autoencoders
\item
  DL - Deep Learning
\item
  MARS - Multivariate Adaptive Regression Splines
\item
  SMD - Standardized Mean Difference
\item
  TMLE - Targeted Maximum Likelihood Estimation
\item
  RHC - Right Heart Catheterization
\item
  SUPPORT - Study to Understand Prognoses and Preferences for Outcomes
  and Risks of Treatments
\end{enumerate}

>>>>

The results for each method under the different scenarios are summarized
below. See Figures \ref{fig:bias-comparison} and
\ref{fig:Coverage-comparison} for an overview of the performance in
terms of bias and coverage, respectively.

\begin{enumerate}
\def\labelenumi{(\roman{enumi})}
\tightlist
\item
  \textbf{Frequent Exposure and Outcome (base) scenario}:
\end{enumerate}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \emph{Bias}: The kitchen sink model, which includes all variables
  without selection, exhibited the smallest bias (0.0002). In contrast,
  the Genetic Algorithm (GA) showed the highest bias (0.0287). Among the
  other methods, Bross (-0.0001), Hybrid (0.0016), and Elasticnet
  (0.0036) demonstrated low bias, indicating that these methods provide
  estimates closer to the true effect. XGBoost (0.0074) and Random
  Forest (RF) (0.0034) had slightly higher bias but remained within
  acceptable limits.
\item
  \emph{Coverage}: Most methods, including Hybrid, Forward, Backward,
  LASSO, and Elasticnet, achieved high coverage values around 98\(\%\),
  indicating well-calibrated confidence intervals. However, the GA
  method had significantly lower coverage (83.8\(\%\)), suggesting that
  its confidence intervals might be too narrow or biased, potentially
  missing the true effect. Given the bias in the GA method's results, we
  also calculated bias-eliminated coverage. This adjustment improved
  GA's coverage to 96\(\%\), but it still remained lower than that of
  the other methods.
\item
  \emph{Mean Squared Error (MSE)}: XGBoost achieved the lowest MSE
  (0.0006), indicating it as the most accurate method overall.
  Conversely, the GA method had the highest MSE (0.0016), reflecting its
  higher bias and variability. The kitchen sink model (0.0009), Bross
  (0.0008), Hybrid (0.0008), and Elasticnet (0.0009) all had relatively
  similar and moderate MSE values.
\item
  \emph{Standard Error (SE)}: The lowest Empirical SE was observed with
  XGBoost (0.0229), indicating high precision in its estimates. The
  kitchen sink model had the highest Empirical SE (0.0305), suggesting
  greater variability. Other methods such as GA (0.0274), Hybrid
  (0.0278), and Bross (0.0287) exhibited moderate variability, while
  LASSO (0.0299) and Elasticnet (0.0294) had slightly higher
  variability. Similarly, XGBoost also showed the lowest Model-based SE
  (0.0268), consistent with its low Empirical SE and indicating high
  precision. The kitchen sink model had the highest Model-based SE
  (0.0333), suggesting less precision in its estimates.
\end{enumerate}

\textbf{(ii) Rare Exposure and Frequent Outcome}:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \emph{Bias}: In this scenario, the kitchen sink model showed a
  relatively low bias (0.0025), but the Genetic Algorithm (GA) had the
  highest bias (0.0408), indicating a substantial deviation from the
  true effect. XGBoost (0.0259) also exhibited higher bias compared to
  other methods. On the other hand, Bross (0.0035), Hybrid (0.0049), and
  Elasticnet (0.0053) demonstrated moderate bias, while Random Forest
  (RF) (0.0127) and Forward Selection (0.0108) had slightly higher bias
  but remained within an acceptable range.
\item
  \emph{Coverage}: Most methods maintained high coverage levels, with
  XGBoost showing the highest coverage (96.2\(\%\)), suggesting that its
  confidence intervals were well-calibrated despite the higher bias. The
  GA method, however, had slightly lower coverage (92.2\(\%\)),
  indicating that its confidence intervals might be narrower,
  potentially excluding the true effect. Other methods such as RF,
  Forward, Backward, and Hybrid had coverage values around 95\(\%\),
  suggesting adequate interval calibration. To account for the bias in
  GA, bias-eliminated coverage was calculated, which improved the
  coverage for GA to 94.2\(\%\), still slightly lower than other
  methods.
\item
  \emph{Mean Squared Error (MSE)}: XGBoost and Hybrid methods both
  demonstrated the lowest MSE (0.0032), indicating that these methods
  were the most accurate in this scenario. The GA method, with a higher
  MSE (0.0043), reflected its substantial bias and variability. The
  kitchen sink model had an MSE of 0.0043, similar to GA, while other
  methods such as Bross, RF, and Elasticnet exhibited moderate MSE
  values, indicating reasonable accuracy.
\item
  \emph{Standard Error (SE)}: The lowest Empirical SE was observed with
  XGBoost (0.0507) and GA (0.0510), reflecting the high precision of
  these methods despite the higher bias. The kitchen sink model
  exhibited the highest Empirical SE (0.0656), indicating greater
  variability. Methods like Hybrid (0.0564), Bross (0.0595), and RF
  (0.0609) showed moderate variability, while Elasticnet (0.0597) and
  LASSO (0.0587) had slightly higher variability. In terms of
  Model-based SE, XGBoost (0.0531) and GA (0.0533) also showed low
  variability, while the kitchen sink model had the highest Model-based
  SE (0.0623), suggesting less precision in its estimates.
\end{enumerate}

\textbf{(iii) Frequent Exposure and Rare Outcome}:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \emph{Bias}: In this scenario, the kitchen sink model exhibited a
  moderate negative bias (-0.0093), similar to the Bross method
  (-0.0088). The Genetic Algorithm (GA) showed a significantly higher
  bias (0.0362), indicating a substantial deviation from the true
  effect. Among other methods, XGBoost demonstrated the lowest bias
  (-0.0061), while methods like Hybrid (-0.0082), Forward (-0.0070), and
  Backward (-0.0070) had slightly higher but still moderate biases.
  Elasticnet and LASSO both had biases of -0.0079, reflecting slightly
  larger deviations compared to XGBoost but still within acceptable
  limits.
\item
  \emph{Coverage}: Most methods achieved good coverage, with XGBoost,
  Random Forest (RF), and Forward Selection each achieving a coverage
  rate of 95.4\(\%\), indicating well-calibrated confidence intervals.
  The GA method, however, had slightly lower coverage (91.8\(\%\)),
  indicating that its confidence intervals might be narrower,
  potentially excluding the true effect. Bross and the kitchen sink
  model had slightly lower coverage values of 93.8\(\%\) and 93.4\(\%\),
  respectively. After accounting for bias, the bias-eliminated coverage
  for most methods, except GA, remained high, with values ranging from
  98.4\(\%\) to 99.0\(\%\), indicating that most methods effectively
  adjusted for bias in their coverage estimates. GA's bias-eliminated
  coverage was lower at 93.4\(\%\), reflecting its higher inherent bias.
\item
  \emph{Mean Squared Error (MSE)}: XGBoost exhibited the lowest MSE
  (0.0003), indicating it as the most accurate method overall in this
  scenario. GA had the highest MSE (0.0040), reflecting its substantial
  bias and variability. The kitchen sink model (0.0005), Bross (0.0005),
  and other methods like Hybrid (0.0004) and Elasticnet (0.0005) all had
  relatively similar MSE values, indicating moderate accuracy.
\item
  \emph{Standard Error (SE)}: The lowest Empirical SE was observed with
  XGBoost (0.0152), reflecting high precision in its estimates. The GA
  method exhibited the highest Empirical SE (0.0523), indicating greater
  variability and less precision. Methods like Hybrid (0.0184), Forward
  (0.0187), and Elasticnet (0.0203) showed moderate variability, while
  Bross (0.0206) and the kitchen sink model (0.0212) had slightly higher
  variability. In terms of Model-based SE, XGBoost (0.0179) again showed
  the lowest variability, consistent with its low Empirical SE,
  indicating that it provided the most stable estimates. The kitchen
  sink model had a slightly higher Model-based SE (0.0219), indicating
  less precision in its estimates.
\end{enumerate}

@@

\caption{Comparison of Bias Across Different Methods in hdPS Analysis\label{fig:bias-comparison}}\label{fig:unnamed-chunk-1}
\end{figure}

@@

\caption{Comparison of Coverage Probability Across Different Methods in hdPS Analysis\label{fig:Coverage-comparison}}\label{fig:unnamed-chunk-2}
\end{figure}

@@

See Figure \ref{fig:comparison-plot} for the results from analyzing the
NHANES (2013-2018) dataset.

@@

\caption{Figure presenting a comparison of Risk Differences (RD) and Odds Ratios (OR) with 95\% confidence intervals for different methods used to evaluate the association between obesity and diabetes risk. The analysis is based on data from the National Health and Nutrition Examination Survey (NHANES) for the years 2013-2018. Methods are arranged by the number of variables used in the models.\label{fig:comparison-plot}}\label{fig:unnamed-chunk-3}
\end{figure}

@@

\begin{itemize}
\tightlist
\item
  hdPS: High-dimensional Propensity Score
\item
  NHANES: National Health and Nutrition Examination Survey
\item
  OR: Odds Ratio
\item
  RD: Risk Difference
\item
  SE: Standard Error
\item
  MSE: Mean Squared Error
\item
  LASSO: Least Absolute Shrinkage and Selection Operator
\item
  GA: Genetic Algorithm
\item
  RF: Random Forest
\item
  XGBoost: Extreme Gradient Boosting
\item
  FS: Forward Selection
\item
  BE: Backward Elimination
\item
  CV: Cross-Validation
\item
  RR: Relative Risk
\item
  Elasticnet: A regularized regression method that combines LASSO and
  Ridge regression
\end{itemize}
